A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-10-19 thesis/thesis.tex
finalize introductio

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{invariant}[theorem]{Invariant}

@@

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress. For the reasons told, we are interested in the design of the data structures that do not use locks.

The question that may arise is ``What properties matter for a lock-free data structure?''. Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty. The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}. The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution that is not linearizable. Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or \nf{null}.}
\end{figure}

@@

\it{is this place good for queue definition?}
A queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored. \texttt{Dequeue()} removes and returns the first element among in the sequence. If the queue is empty it returns \texttt{null}.

\paragraph{Model} 
A system is called \it{asynchronous} when processes in the system run at arbitrary varying speeds, i.e the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy showed \cite{10.1145/114005.102808} that we one cannot implement any lock-free data structures with multi-reader multi-writer registers alone. In this work we use Compare \& Swap (\nf{CAS}) objects to reach universal synchronization.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related word done in the area and the motivation of our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2 we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. Section 3.2 contains the algorithm itself.

We prove the correctness of our queue in Chapter 4 by showing it is wait-free and linearizable.

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.

Finally, we talk about some concluding remarks in Chapter 6. It contains how we can improve our algorithm memory usage and how to use its idea in designing other wait-free data structures in section 6.

\pagebreak
\section{Related Work}
In this Section, we look at previous lock-free queues.
\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). A shared variable Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

@@

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queues. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node. They change the direction of storing the nodes and use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As previous algorithms, the worst case happens when contenion is high: $p$ concurrent enqueues have happened, which they have to be appended to the linked list one by one. The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.

Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining. Their queue is linearizable but not lock-free. Their main idea is that with knowledge of all the history of operations, it might be possible to answer queries faster than doing them one by one. In our work we also maintain the whole history. They present experiments that show their algorithm performs well in some situations.

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. There is a global head which points to arrays containing the first elements in the queue. Also there is a head that points to the first element in the array that global head points. Global tail and tail pointers are the same. Global pointers are up to date, but head and tail pointers may be behind in time. An enqueue or a dequeue searches in the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}). Their data structure is lock-free. Still, if the head array is empty and $p$ processes try to enqueue simultaneously, the step complexity remains $\Omega(p)$.

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue with batching. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the size of the queue before and after a batch which is similar to apart of our work. Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a new notion introduced by the authors called Extended Medium Futures Linearizability.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. They design a slow phase which is wait-free bounded. If a process has no progress other processes try to help it in its \nf{CAS} loop in its slow phase. Since there can be $p$ processes helping the worst-case complexity cannot be better than linear.

In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case where all $p$ processes try to do an enqueue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}. It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09}. Our motivation is to overcome this problem and present a wait-free sublinear queue.

@@

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. Their algorithm time complexity is $O(\log n)$ where $n$ is the number of enqueuers. We will improve their algorithm to construct a multi-enqueuer multi-dequeuer queue.

\bf{TODO: Where should be?}
Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the list of elements added by processes and when a process wants to do Fetch\&Inc it appends its element to the tree and return the number of elements in the tree. Their work can be descibed a queue without dequeues.

\subsection{Universal Constructions}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects. We mention a non-practical poly-logarithmic universal construction in page 12.

@@

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue. Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.

@@

\section{Analysis}

Is this a Lemma or a Result or an Observation of Algorithm? Which is not used in main wait-free theorem.
\begin{lemma}
An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
\end{lemma}
\begin{proof}
  In each level of the tree  \nf{Refresh} is invoked at most two times and  every \nf{Refresh} invokes at most 7 \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
\end{proof}

@@

\pagebreak
\section{Future Directions}
We designed a tree to achieve agreement on an linearization of operations invoked by $p$ processes in an asynchronous model, which we will call \it{Block Tree}. We also implemented two queries to know information about the ordering agreed in the block tree. Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes. Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects. In the next paragraphs we talk about possible improvements on block trees and the data structures that we can implement with block trees.

\paragraph{Reducing Space Usage}
The \nf{blocks} arrays defined in our algorithm are unbounded. We could instead use the vector of Feldman, Valera-Leon and  Damian Segment~\cite{7073592} model to use $O(n)$ space in each node, where $n$ is the total number of operations. Their vector creates an array called \nf{arr} of pointers to array segments. When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not. If not, it creates a shared array with size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}. Whether the \nf{CAS} is successful or not \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array. When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps. Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}. Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]}. Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps. The time taken to allocate arrays in an execution containing $n$ operations is $O(p\log n)$ which if $n>>p$ we can ignore the $p$ factor. The vector implementation has also a mechanism for doubling \nf{arr} when necessary, but this happens very rarely, since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
\begin{figure}[hbt]  
  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
  \caption{Array Segments}
  \label{fig::doublingArray}
\end{figure}

\paragraph{Garbage Collection}
We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e, all of its enqueues have been dequeued and the responses of all of its dequeues have been computed. We call these blocks \it{finished blocks}. If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished then all blocks before $B$ are also finished. Knowing the most recent finished block in a node we can reclaim the memory taken by finished blocks. To throw the garbage in the blocks away we cannot use arrays (or vectors). We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$. Where \nf{split(i)} removes all the indices less than \nf{i}. We can use a concurrent implementation of persistent red black trees for this~\cite{DBLP:conf/podc/BashariW21}.

@@

\paragraph{Poly-logarithmic Wait-free Sequences}
Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}. An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence. We can modify our queue to design such data structure. \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} with replacing the \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
 We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.

\paragraph{Stacks}
There are two reasons block tree worked well to implement a queue. Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return\nf{null} we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size. Secondly, the operations we need to search to respond to the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue. It may be possible to create wait-free poly-logarithmic implementaion of other objects whose operations satisfy these two conditions.

\paragraph{Attiya--Fouren Lower Bound}
Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes). Their result says any implementation of queues, using reads,
writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$
amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.

@@

\bibliography{main.bib} 
\bibliographystyle{plain}

>>>>

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{invariant}[theorem]{Invariant}
\newtheorem{proposition}[theorem]{Proposition}

@@

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure to work with. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty it returns \texttt{null}. In a concurrent version of queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrary varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy showed \cite{10.1145/114005.102808} that we one cannot implement concurrent version of all data structures with multi-reader multi-writer registers alone, introducing a hieachy on how powerful are objects to solve consensus problem among processes. Objects like LL/SC, CAS can be used to reach consensus among any number of processors. CAS object implements an atomic Compare \& Swap(\nf{new, old}) operation, which if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns {true}, otherwise returns \nf{false}.  In this work we use Compare \& Swap (\nf{CAS}) objects to reach universal synchronization.

\bf{Concurrent execution: I didn't understed interleaving steps of processes}

The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty. The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}. The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can be occurred before termination of \texttt{Enqueue(A)} and return \nf{null} or can be occurred after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

@@

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free bounded poly-log time queue. We design a tournament tree in which processes try to propagate their operations up to the root to be linearized and they can do queries to get information about the linearization stored in the root. To solve the CAS Retry problem when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for processor. Each operation in our queue does at most $14\log p$ \nf{CAS} operations. We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations. The garbage collection remains open which we have pointed to a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related word done in the area and the motivation of our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2 we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 is summoned with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also talk about the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself.

We prove the correctness of our queue in Chapter 4 by showing it is linearizable.

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue. In the end we prove our queue is wait-free.

Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues.
\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). A shared variable Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

@@

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. Asin  previous algorithms, the worst case happens in the case where the contention is high when $p$ concurrent enqueues happene, their nodes have to be appended to the linked list one by one. The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.

Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining. The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure. Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is occupied an empty array is linked to the array and tail pointers are updated. Global head points to the array containing the first element in the queue, and head that points to the first element in that array. Global tail and tail pointers are similar (see~Figure~\ref{fig::sundell}). The pointers might be behind their real places, a process updates the position of the pointers when using them. When a process wants to write an element to current head it uses \nf{CAS}, so if $p$ processes try to enqueue simultaneously, the step (also \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supprting futures. A future is an operation that can be applied on the queue when it is requested. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the size of the queue before and after a batch which is similar to a part of our work. Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a safety condition weaker than linearizabilty introduced by the authors, called Extended Medium Futures Linearizability.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. They design a slow phase which is wait-free bounded. If a process has no progress other processes try to help it in its \nf{CAS} loop in its slow phase. Since there can be $p$ processes helping, the worst-case complexity cannot be better than $\Omega(p)$.

In the worst-case, step complexity of all the list-based queues discussed above, includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}. It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}. Our motivation is to overcome this problem and present a wait-free sublinear queue.

@@

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a poly-logarithmic multi-enqueuer multi-dequeuer queue. Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.

\subsection{Universal Constructions and other Poly-log Time Data Structures}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log n)$ worst case step complexity, using atomic access to $\Omega(n \log n)$ bit words. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects. We mention a non-practical universal construction with poly-log \nf{CAS} invocations in page 12.

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes and when a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree. If we have an array and two Fetch\&Inc objects head and tail we can implement a queue: to do an enqueue we can Fetch\&Inc head and write in the value fetched of the array and to do a dequeue we can Fetch\&Inc the tail and fetched the array's cell in the value returned.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list based queues one might guess $p$ factor is inherent in time complexity of concurrent queues. Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes). Their result says any implementation of queues, using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that their result does not apply on our design and we manage to reach poly-log time complexity. It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.

@@

@@

\section{Analysis}
In this section we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
\begin{proposition}
An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
\end{proposition}
\begin{proof}
  In each level of the tree  \nf{Refresh} is invoked at most two times and  every \nf{Refresh} invokes at most 7 \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
\end{proof}

@@

\pagebreak
\section{Future Directions}
We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}. We  implemented two queries to know information about the ordering agreed in the block tree. Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes. Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects. In the next paragraphs we talk about possible improvements on block trees and the data structures that we can implement with block trees.

\paragraph{Reducing Space Usage}
The \nf{blocks} arrays defined in our algorithm are unbounded. To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}. We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}). When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not. If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}. Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array. When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps. The CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}. Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]}. Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps. The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$. The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely, since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
\begin{figure}[hbt]  
  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
  \caption{Array Segments}
  \label{fig::doublingArray}
\end{figure}

\paragraph{Garbage Collection}
We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e, all of its enqueues have been dequeued and the responses of all of its dequeues have been computed. We call these blocks \it{finished blocks}. If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished then all blocks before $B$ are also finished. Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks. To throw the garbage in the blocks away we cannot use arrays (or vectors). We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$, where \nf{split(i)} removes all the indices less than \nf{i}. If each process tries to do the garbage collection every $p^2$ operation on the queue then the amortized complexity remains the same. We can use a concurrent implementation of persistent red black trees for this~\cite{DBLP:conf/afp/Okasaki96}. Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red black trees in a similar way.

@@

\paragraph{Poly-logarithmic Wait-free Sequences}
Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}. An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence. We can modify our queue to design such data structure. An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
 We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.

\paragraph{Stacks}
There are two reasons the block tree worked well to implement a queue. Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size. Secondly, the operations we need to search to respond to the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue. It may be possible to create wait-free poly-logarithmic implementaion of other objects whose operations satisfy these two conditions.

@@

\bibliography{main.bib} 
\bibliographystyle{acm}
